---
title: "Getting Started with MID Collections"
description: |
  Overview of the collection objects: midlist and midrib.
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
pkgs <- c("ggplot2", "patchwork", "gam", "ranger", "ISLR2", "viridisLite")
for (pkg in pkgs) {
  if (!(pkg %in% installed.packages())) {
    install.packages(pkg)
  }
}
```

The **midr** package is designed not only to interpret individual models but also to facilitate **comparative analysis**. When you are dealing with multiple outcomes or comparing different modeling approaches, **midr** provides two specialized collection classes: "midlist" and "midrib".

These collections allow you to visualize and compare feature effects, importance, and instance-level breakdowns across multiple models using a unified interface.

## midlist: Flexible Combination of MID Objects

The "midlist" class is a versatile container used to group existing MID objects (such as `mid`, `midimp`, `midbrk`, or `midcon`). This is particularly useful when you have trained separate models—perhaps using different algorithms or targeting different subsets of data—and want to compare their behavior side-by-side.

### Example: Comparing GAM vs. Random Forest

When predicting the total number of users (`bikers`) in the Bikeshare dataset, different machine learning algorithms capture non-linear relationships differently.
Here, we fit a Generalized Additive Model (**gam**) and a Random Forest (**ranger**), interpret both of them independently, and unify the results into a single "midlist" collection for a direct visual comparison.

```{r midlist-demo, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
library(midr)
library(gam)
library(ranger)

data(Bikeshare, package = "ISLR2")

fit.gam <- gam(
  bikers ~ mnth + hr + s(weekday) + weathersit + s(temp) + s(hum),
  data = Bikeshare
)

fit.ranger <- ranger(
  bikers ~ mnth + hr + weekday + weathersit + temp + hum,
  data = Bikeshare
)

# Interpret two separate models independently
mid.gam <- interpret(
  bikers ~ mnth + hr + weekday + weathersit + temp + hum,
  data = Bikeshare,
  model = fit.gam,
  k = 50
)

mid.ranger <- interpret(
  bikers ~ mnth + hr + weekday + weathersit + temp + hum,
  data = Bikeshare,
  model = fit.ranger,
  k = 50,
  lambda = 1
)

# Combine them into a midlist
mids <- midlist(
  gam = mid.gam,
  ranger = mid.ranger
)
class(mids)
```

When `ggmid()` is called on a "midlist" collection, it automatically handles the comparison.

```{r midlist-plot}
options(midr.qualitative = "viridis")

p1 <- ggmid(mids, "hr") + theme(legend.position = "bottom")
p2 <- ggmid(mids, "temp") + theme(legend.position = "none")
p3 <- ggmid(mids, "hum") + theme(legend.position = "none")

p1 / (p2 + p3)
```

**Mathematical Note:** The "midlist" architecture offers maximum flexibility.
Each model in the collection can maintain its own unique fitting parameters such as `lambda` (regularization strength), `k` (number of knots), and `type` (shape of component functions).

## midrib: Efficient Multi-Response MID Model

The "midrib" class is designed for **multi-output scenarios**, where a single model predicts multiple target variables simultaneously. Instead of interpreting each response in isolation, **midr** treats the multi-output structure as a single entity—a "midrib" or shared backbone.

You can trigger the creation of a "midrib" object in two ways:

1. **via formula**: Provide a matrix or data frame as the response (e.g., using `cbind()`).
2. **via prediction function**: Pass a model and a custom `pred.fun` that returns a matrix or data frame of predictions.

### Example: Joint Interpretation of Registered and Casual Users

Instead of comparing different algorithms, we might want to understand how features simultaneously affect different segments of our target variable.
Here, we interpret a single model predicting both "registered" and "casual" users.

```{r midrib-demo}
# Using a formula with a multi-column response
midrib <- interpret(
  cbind(registered, casual) ~ (mnth + hr + workingday + weathersit + temp + hum),
  data = Bikeshare,
  k = 50,
  lambda = 1
)

class(midrib)
```

The resulting "midrib" object stores the fitted functions (i.e., all coefficients) for all outcomes in a unified structure, allowing for seamless comparative plotting.

```{r midrib-plot}
options(midr.qualitative = "cividis")

p1 <- ggmid(midrib, "hr") + theme(legend.position = "bottom")
p2 <- ggmid(midrib, "temp") + theme(legend.position = "none")
p3 <- ggmid(midrib, "hum") + theme(legend.position = "none")

p1 / (p2 + p3)
```

**Mathematical Note:** By sharing a single design matrix across all responses, the "midrib" class significantly reduces memory consumption and computation time.
This makes it the ideal choice for high-dimensional multi-output data.


## Summary of Collection Structures

| Class | Data Structure | Optimization Logic | Key Advantage |
| :--- | :--- | :--- | :--- |
| **"midlist"** | A list of "mid", "midimp", "midbrk", or "midcon". | **Independent**: Each model has its own fitting parameters. | **Flexibility**: Compare heterogeneous models. |
| **"midrib"** | A single multivariate response model. | **Joint**: Shares a single design matrix across all responses. | **Efficiency**: Significant speedup for multivariate targets. |
