---
title: "Getting Started with Color Themes"
output: rmarkdown::html_vignette
description: >
  Provides a gallery of color themes implemented in the midr package.
vignette: >
  %\VignetteIndexEntry{Getting Started with Color Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 2.2,
  fig.width = 6
)

pkgs <- c("khroma", "viridisLite", "RColorBrewer", "gridExtra", "ggplot2")
for (pkg in pkgs) {
  if (!(pkg %in% installed.packages())) {
    install.packages(pkg)
  }
}
```

The "color.theme" object is designed as a container for two color functions: `palette()` and `ramp()`.
It can be created by specifying a color theme name as the first argument of the `color.theme()` function.

```{r, fig.height = 1.2, fig.width = 6, echo = c(-3, -4)}
library(midr)
library(ggplot2)
theme_set(theme_midr())
par.midr()
library(gridExtra)
nightfall <- color.theme("nightfall")
print(nightfall)
nightfall$palette(3)
nightfall$ramp(c(0.00, 0.25, 0.50, 0.75, 1.00))
accent <- color.theme("Accent", pkg = "grDevices")
print(accent)
accent$palette(5)
viridis <- color.theme("viridis")
print(viridis)
viridis$palette(3)
viridis$ramp(c(0.00, 0.25, 0.50, 0.75, 1.00))
```

Alternatively, a new color theme can be created by passing a color vector.

```{r, fig.height = 1.2, fig.width = 6, echo = -1}
par.midr()
divct <- color.theme(c("violetred", "white", "darkblue"), type = "diverging")
print(divct)
seqct <- color.theme(c("white", "steelblue", "darkblue", "violetred"))
print(seqct)
qualct <- color.theme(c("violetred", "steelblue", "gray50"), type = "qualitative")
print(qualct)
```

The color appearance of visualizations created with **midr** can be easily customized by passing a "color.theme" object or a color theme name implemented in **midr** (see below) to `ggmid()` or `plot()`.

```{r, fig.weight = 6, fig.height = 4}
set.seed(42)
dataset <- diamonds[sample(nrow(diamonds), 5000L), ]
mid <- interpret(price ~ (carat + color + clarity + cut) ^ 2, dataset)
grid.arrange(
  ggmid(mid, "color:clarity", main.effect = TRUE),
  ggmid(mid, "color:clarity", main.effect = TRUE, theme = divct),
  ggmid(mid, "carat:color", main.effect = TRUE, theme = "viridis"),
  ggmid(mid, "carat:color", main.effect = TRUE, theme = "viridis_r")
)
imp <- mid.importance(mid)
grid.arrange(
  ggmid(imp, "heatmap"),
  ggmid(imp, "barplot", max = 10, theme = qualct),
  ggmid(imp, "heatmap", theme = seqct),
  ggmid(imp, "barplot", max = 10, theme = "dark")
)
```

The functions `scale_color_theme()` and `scale_fill_theme()` provide a pipeline for customizing ggplot graphics.

```{r, fig.weight = 6, fig.height = 4}
p <- ggplot(dataset) + geom_point(aes(carat, price, col = color))
grid.arrange(
  p + scale_color_theme("DALEX"),
  p + scale_color_theme("discreterainbow"),
  p + scale_color_theme("Okabe-Ito"),
  p + scale_color_theme("plasma", discrete = TRUE)
)
```

```{r sequential, echo = FALSE}

themes <- list()

themes[[1L]] <- data.frame(
  pack = "viridisLite",
  name = c("magma", "inferno", "plasma", "viridis",
            "cividis", "rocket", "mako", "turbo"),
  type = "s"
)

temp <- RColorBrewer::brewer.pal.info
themes[[2L]] <- data.frame(
  pack = "RColorBrewer",
  name = rownames(temp),
  type = substr(temp$category, 1L, 1L)
)

temp <- khroma::info()
themes[[3L]] <- data.frame(
  pack = "khroma",
  name = temp$palette,
  type = substr(temp$type, 1L, 1L)
)

themes[[4L]] <- data.frame(
  pack = "grDevices",
  name = hcl.pals(type = "divergingx"),
  type = "d"
)

themes[[5L]] <- data.frame(
  pack = "grDevices",
  name = hcl.pals(type = "diverging"),
  type = "d"
)

themes[[6L]] <- data.frame(
  pack = "grDevices",
  name = hcl.pals(type = "sequential"),
  type = "s"
)

themes[[7L]] <- data.frame(
  pack = "grDevices",
  name = hcl.pals(type = "qualitative"),
  type = "q"
)

themes[[8L]] <- data.frame(
  pack = "grDevices",
  name = grDevices::palette.pals(),
  type = "q"
)

themes[[9L]] <- data.frame(
  pack = "midr",
  name = c("midr", "DALEX", "shap", "grayscale", "bluescale"),
  type = c("d", "q", "s", "s", "s")
)

.df <- unique(do.call(rbind, themes))
.df <- .df[order(.df$type, .df$name, .df$pack),]
```

## Diverging Color Themes

```{r, echo = FALSE}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "d", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 2L], pkg = df[i, 1L])
  text <- sprintf("%s (%s)", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

## Qualitative Color Themes

```{r, echo = FALSE}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "q", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 2L], pkg = df[i, 1L])
  text <- sprintf("%s (%s)", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

## Sequential Color Themes

```{r, echo = FALSE}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "s", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 2L], pkg = df[i, 1L])
  text <- sprintf("%s (%s)", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

