---
title: "Visualization with ggplot2 package"
description: |
  Tutorial for visualizations with the ggmid() function.
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 9,
  fig.height = 3.5
)
pkgs <- c("ggplot2", "patchwork")
for (pkg in pkgs) {
  if (!(pkg %in% installed.packages())) {
    install.packages(pkg)
  }
}
```

## Introduction

This tutorial explains how to flexibly visualize MID models built with the **midr** package using the **ggplot2** ecosystem.

Since the `ggmid()` function returns a ggplot object, you can easily combine plots using **patchwork** or add advanced customizations via `geom_*()` functions.
Through this article, we will explore step-by-step how to create rich figures suitable for papers and reports, starting from the default plots.

```{r packages, message=FALSE, warning=FALSE}
library(midr)
library(ggplot2)
library(patchwork)
library(survival)
theme_set(theme(legend.position = "bottom"))
.transparent <- "#ffffff00"
```

## Single MID Model

First, let's look at basic visualizations using a single MID model.
Here, we use the built-in `diamonds` dataset to build a model that predicts the diamond's price.

```{r diamonds}
# fit a MID model
mid <- interpret(
  log(price) ~ (carat + cut + color + clarity) ^ 2,
  k = 5L, # number of basis functions
  lump = "auto", # lump factor levels
  data = diamonds,
  verbosity = 0L
)
diamonds_sample <- diamonds[sample(nrow(diamonds), 1000L), ]
mid$call$data <- quote(diamonds_sample)
```

### First-Order Effect (Numeric Variable)

To examine the effect of a continuous variable, specify the variable name in the `term` argument.

```{r, fig.width = 3.5}
p1 <- ggmid(mid, term = "carat")
p1
```

In addition to the default plot, you can also see the underlying data and aesthetic mappings (`aes`) that `ggmid()` generates internally.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

By combining standard ggplot2 functions, you have a high degree of freedom to customize the plot—such as changing line widths and colors, or overlaying actual data points with `geom_point()`.

```{r}
p1 <- ggmid(mid, term = "carat", linewidth = 2, color = "dodgerblue4")
p2 <- ggmid(mid, term = "carat", color = .transparent) +
  geom_point(aes(y = log(price) - mean(log(price))), data = diamonds_sample) +
  geom_line(color = "firebrick1")
p3 <- ggmid(mid, term = "carat", type = "data", theme = "shap") +
  geom_line(linewidth = 4, alpha = .2)
p1 + p2 + p3
```

### First-Order Effect (Factor Variable)

Categorical variables can be visualized in a similar manner.
If the `lump` argument was specified during model fitting, factor levels are automatically grouped together.
However, by setting `lumped = FALSE` when plotting, you can expand and display all levels.

```{r, fig.width = 7}
p1 <- ggmid(mid, term = "clarity")
p2 <- ggmid(mid, term = "clarity", lumped = FALSE)
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

You can also overlay bar charts or use `geom_jitter()` to simultaneously represent the data distribution for each category.

```{r}
p1 <- ggmid(mid, term = "clarity", fill = "dodgerblue4", lumped = FALSE)
p2 <- ggmid(mid, term = "clarity", fill = .transparent, lumped = FALSE) +
  geom_jitter(aes(y = log(price) - mean(log(price))), height = 0, data = diamonds_sample) +
  geom_col(fill = NA, color = "firebrick1")
p3 <- ggmid(mid, term = "clarity", type = "data", theme = "shap", shape = "|") +
  geom_line(aes(group = NA), linewidth = 4, alpha = .2)
p1 + p2 + p3
```

### Second-Order Effect

This visualizes the interaction between two variables (`term = "var1:var2"`).
By specifying `main.effects = TRUE`, you can view the overall effect, including the main effects.

```{r, fig.width = 7}
p1 <- ggmid(mid, term = "carat:clarity")
p2 <- ggmid(mid, term = "carat:clarity", main.effects = TRUE)
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

Additionally, by utilizing heatmap-oriented themes (e.g., `theme = "Heat"`, `"mako"`), complex second-order effects can be represented intuitively.

```{r}
p1 <- ggmid(mid, term = "carat:clarity", type = "compound",
            theme = "Heat", size = 1, shape = 1, lumped = FALSE)
p2 <- ggmid(mid, term = "carat:clarity", type = "data",
            theme = "mako", main.effects = TRUE)
p3 <- ggmid(mid, term = "carat:clarity", type = "data",
            theme = "shap", main.effects = TRUE) +
  geom_line(aes(group = clarity, color = mid), linewidth = 4, alpha = .2)
p1 + p2 + p3
```

### Feature Importance

Here, we visualize the importance of each variable calculated by the `mid.importance()` function.
`ggmid()` supports various plot formats depending on your needs through the type argument.

```{r, fig.width = 7}
imp <- mid.importance(mid)
p1 <- ggmid(imp)
p2 <- ggmid(imp, type = "dotchart")
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(imp, theme = "light")
p2 <- ggmid(imp, type = "dotchart", theme = "shap", size = 3)
p3 <- ggmid(imp, type = "dotchart", color = .transparent) +
  geom_label(aes(x = importance, label = round(importance, 2)), size = 3) +
  lims(x = c(-0.05, 1.05))
p1 + p2 + p3
```

The heatmap style provides a high-level overview of importance values in a heatmap format.

```{r, fig.width = 3.5}
p1 <- ggmid(imp, type = "heatmap")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

Adding text (`geom_text()`) to explicitly show the values makes it even more reader-friendly.

```{r}
p1 <- ggmid(imp, type = "heatmap", fill = "white", color = "black") + 
  geom_text(aes(label = round(importance, 2)))
p2 <- ggmid(imp, type = "heatmap", theme = "mako_r") +
  geom_text(aes(label = round(importance, 2)), color = "white")
p3 <- ggmid(imp, type = "heatmap", fill = .transparent) +
  geom_point(aes(size = importance), color = "steelblue")
p1 + p2 + p3
```

The boxplot style represents the distribution of importance in detail.

```{r, fig.width = 3.5}
p1 <- ggmid(imp, type = "boxplot")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(imp, type = "boxplot", theme = "mako")
p2 <- ggmid(imp, type = "boxplot", fill = .transparent, color = .transparent) +
  geom_jitter(size = .5, width = 0) +
  geom_boxplot(color = "firebrick1", fill = NA)
p3 <- ggmid(imp, type = "boxplot", fill = .transparent, color = .transparent) +
  geom_violin(aes(fill = importance), scale = "width")
p1 + p2 + p3
```

### Prediction Breakdown

Using `mid.breakdown()`, we decompose and display the contribution of each feature to the prediction for a specific observation.

```{r, fig.width = 3.5}
brk <- mid.breakdown(mid, row = 1L)
p1 <- ggmid(brk)
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(brk, theme = "shap@qual", color = .transparent, width = .95) +
  geom_text(aes(label = round(mid, 2), x = (xmin + xmax) / 2), size = 3)
p2 <- ggmid(brk, pattern = c("%t", "%t, %t"), width = .05)
p3 <- ggmid(brk, pattern = c("%t\n%v", "%t:%t\n%v:%v"), max.nterms = 7)
p1 + p2 + p3
```

You can flexibly adjust the layout using the pattern and type arguments, ranging from waterfall-like charts to bar plots and dot charts.

```{r, fig.width = 7}
p1 <- ggmid(brk, type = "barplot")
p2 <- ggmid(brk, type = "dotchart")
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(brk, type = "barplot", theme = "light")
p2 <- ggmid(brk, type = "barplot", theme = "bicolor_r", vline = FALSE) +
  aes(x = abs(mid)) +
  geom_text(aes(label = round(mid, 2), x = abs(mid) + 0.03), size = 3)
p3 <- ggmid(brk, type = "dotchart", theme = "shap", size = 3)
p1 + p2 + p3
```

### Conditional Expectation

Using `mid.conditional()`, we can observe how the predicted values transition as a specific variable changes (similar to ICE plots or PDPs).
By specifying `type = "centered"`, you can align the baselines by showing the amount of change from a specific reference point (reference), making it easier to compare the model's behavior across different groups.

```{r, fig.width = 7}
con <- mid.conditional(mid, variable = "carat", data = diamonds_sample[1:100, ])
p1 <- ggmid(con)
p2 <- ggmid(con, type = "centered")
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p2[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(con, points = FALSE, color = "dodgerblue4") +
  geom_point()
p2 <- ggmid(con, type = "centered", var.color = cut,
            shape = 1, size = 3, var.linetype = cut)
p3 <- ggmid(con, type = "centered", theme = "light",
            var.color = "cut", reference = 50)
p1 + p2 + p3
```

## Collection of MID Models

Next, we will explore visualizations for a collection (set) of multiple MID models.
As an example, we build a Cox proportional hazards model using the `veteran` dataset from the **survival** package, and evaluate the transition of effects over time as a collection of models.

```{r}
# fit a survival mid model
sreg <- coxph(
  Surv(time, status) ~ .,
  data = veteran
)
mids <- interpret(
  Surv(time, status) ~ .,
  data = veteran,
  model = sreg,
  lambda = 1
)
```

### First-Order Effect (Numeric Variable)

```{r, fig.width = 3.5}
p1 <- ggmid(mids, term = "karno")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

When you have multiple models with numeric labels, specifying `type = "series"` allows you to easily create a time-series-like transition graph with the model index (here, survival time) on the x-axis.

```{r, fig.width = 3.5}
p1 <- ggmid(mids, term = "karno", type = "series")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

You can also plot the baseline intercept transition by setting `intercept = TRUE`, or adjust the format for publication using theme functions, such as monochrome (`theme = "grayscale"`).

```{r}
p1 <- ggmid(mids, term = "karno", intercept = TRUE)
p2 <- ggmid(mids, term = "karno", type = "series", intercept = TRUE,
            theme = "mako")
p3 <- ggmid(mids, term = "karno", type = "series", intercept = TRUE,
            theme = "grayscale") +
  geom_line(
    data = data.frame(
      mid = mids$intercept,
      label = as.numeric(labels(mids))
    ),
    color = "firebrick1"
  )
p1 + p2 + p3
```

### First-Order Effect (Factor Variable)

```{r, fig.width = 3.5}
p1 <- ggmid(mids, "celltype")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r, fig.width = 3.5}
p1 <- ggmid(mids, "celltype", type = "series")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r}
p1 <- ggmid(mids, term = "celltype", intercept = TRUE)
p2 <- ggmid(mids, term = "celltype", type = "series", intercept = TRUE,
            theme = "mako")
p3 <- ggmid(mids, term = "celltype", type = "series", intercept = TRUE) +
  geom_line(
    data = data.frame(
      mid = mids$intercept,
      label = as.numeric(labels(mids))
    ),
    linetype = "dotted"
  )
p1 + p2 + p3
```

### Feature Importance

The feature importance for the model collection can also be visualized along the progression of time (or model differences).

```{r, fig.width = 3.5}
imps <- mid.importance(mids)
p1 <- ggmid(imps)
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r, fig.width = 3.5}
p1 <- ggmid(imps, type = "series")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

By using stacked area charts with `geom_area()` or representing distributions with `geom_boxplot()` or `geom_jitter()`, you can visually capture the dynamics of which features become important at what specific timing.

```{r}
p1 <- ggmid(imps, fill = .transparent) +
  geom_boxplot()
p2 <- ggmid(imps, fill = .transparent) +
  geom_jitter(aes(color = label), width = 0) +
  scale_color_theme("mako")
p3 <- ggmid(imps, type = "series", color = .transparent) +
  geom_area(aes(fill = term), position = "fill") +
  scale_fill_theme("light")

p1 + p2 + p3
```

### Prediction Breakdown

Similar to a single breakdown, this visualizes how the contribution of features for a specific observation changes across the entire collection.
It is highly effective for tracking how the local impact of each variable fluctuates over time.

```{r, fig.width = 3.5}
brks <- mid.breakdown(mids, row = 42)
p1 <- ggmid(brks)
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```

```{r, fig.width = 3.5}
p1 <- ggmid(brks, type = "series")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```


```{r}
p1 <- ggmid(brks, fill = .transparent) +
  geom_boxplot()
p2 <- ggmid(brks, fill = .transparent) +
  geom_jitter(aes(color = label), width = 0) +
  scale_color_theme("mako")
p3 <- ggmid(brks, type = "series", color = .transparent) +
  geom_area(aes(fill = term)) +
  scale_fill_theme("light")

p1 + p2 + p3
```

### Conditional Expectation

Finally, let's look at the transition of conditional expectations across multiple models or time axes.
By utilizing `facet_grid(~ .id)`, you can split the panels by time (the model's index) and compare side-by-side how the shape of the expectation evolves.

```{r, fig.width = 7}
cons <- mid.conditional(mids, variable = "karno",
                        max.nsamples = 3L, data = veteran[1:3, ])
p1 <- ggmid(cons)
p2 <- ggmid(cons, type = "centered")
p1 + p2
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p2[[x]], 3)); cat("\n")}
) -> .
```

```{r}
ggmid(cons, theme = "mako", type = "centered", reference = 50) +
  facet_grid(~ .id)
```

```{r, fig.width = 3.5}
p1 <- ggmid(cons, type = "series")
p1
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
lapply(
  c("data", "mapping"),
  function(x) {cat("$", x, "\n", sep = ""); print(head(p1[[x]], 3)); cat("\n")}
) -> .
```


```{r}
ggmid(cons, theme = "mako", type = "series") +
  facet_grid(~ .id)
```
