---
title: "Interpretation of Regression Models"
description: |
  Demonstrates how to use the midr package to interpret regression models.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkgs <- c("mlbench", "gridExtra", "ggplot2", "Metrics", "khroma", 
          "glmnet", "ranger", "e1071", "nnet", "rpart", "gam", "xgboost")
for (pkg in pkgs) {
  if (!(pkg %in% installed.packages())) {
    install.packages(pkg)
  }
}
```

This article presents some examples of the interpretation of regression models using `midr`.

```{r setup, message = FALSE}
# load required packages
library(midr)
library(ggplot2)
library(gridExtra)
library(patchwork)
library(Metrics)
theme_set(theme_midr())
```

## Regression Task

We use a benchmark regression task, originally described in Friedman (1991) and Breiman (1996), and implemented in the `mlbench` package. The dataset has 10 independent predictor variables $x_1,x_2,...,x_{10}$ each uniformly distributed on the interval $[0,1]$, and the response variable $y$ , generated according to the following formula with disturbance term $\epsilon\ {\sim}\ \mathcal{N}{(0, 1)}$. $$
y=10\sin{(\pi{x_1}{x_2})+20{(x_3-0.5)}^2+10x_4+5x_5+\epsilon}
$$

The following plots show the effect of each predictor variable on the response. For $x_1$ and $x_2$ , the interaction effect is shown by the colored lines: the effect of $x_1$ depends on the value of $x_2$ (pale purple for 0 and dark red for 1) and *vice versa*.

```{r dataviz, echo = FALSE}
# visualization of the component functions
f1 <- function(y) {function(x) 10 * sin(pi * x * y)}
f3 <- function(x) 20 * (x - 0.5)^2
f4 <- function(x) 10 * x
f5 <- function(x) 5 * x
cols <- color.theme("smoothrainbow")$ramp
pl <- ggplot() + xlim(0, 1) + xlab("x1, x2") +
  geom_function(fun = f1(0.0), col = cols(0.0)) +
  geom_function(fun = f1(0.1), col = cols(0.1)) +
  geom_function(fun = f1(0.2), col = cols(0.2)) +
  geom_function(fun = f1(0.3), col = cols(0.3)) +
  geom_function(fun = f1(0.4), col = cols(0.4)) +
  geom_function(fun = f1(0.5), col = cols(0.5)) +
  geom_function(fun = f1(0.6), col = cols(0.6)) +
  geom_function(fun = f1(0.7), col = cols(0.7)) +
  geom_function(fun = f1(0.8), col = cols(0.8)) +
  geom_function(fun = f1(0.9), col = cols(0.9)) +
  geom_function(fun = f1(0.05), col = cols(0.05)) +
  geom_function(fun = f1(0.15), col = cols(0.15)) +
  geom_function(fun = f1(0.25), col = cols(0.25)) +
  geom_function(fun = f1(0.35), col = cols(0.35)) +
  geom_function(fun = f1(0.45), col = cols(0.45)) +
  geom_function(fun = f1(0.55), col = cols(0.55)) +
  geom_function(fun = f1(0.65), col = cols(0.65)) +
  geom_function(fun = f1(0.75), col = cols(0.75)) +
  geom_function(fun = f1(0.85), col = cols(0.85)) +
  geom_function(fun = f1(0.95), col = cols(0.95)) +
  geom_function(fun = f1(1.0), col = cols(1.0))
grid.arrange(nrow = 2,
  pl,
  ggplot() + xlim(0, 1) + xlab("x3") + geom_function(fun = f3),
  ggplot() + xlim(0, 1) + xlab("x4") + geom_function(fun = f4),
  ggplot() + xlim(0, 1) + xlab("x5") + geom_function(fun = f5)  
)
```

```{r mlbench}
# benchmark regression task
library(mlbench)
set.seed(42)
train  <- as.data.frame(mlbench.friedman1(n = 500L))
test   <- as.data.frame(mlbench.friedman1(n = 500L))
mtrain <- as.data.frame(mlbench.friedman1(n = 2500L))[, -11L]
```

For each model type, we fit a regression model using the `train` data of 500 observations and an interpretative MID surrogate of the target model using the `mtrain` data of 2500 observations without the response variable. We then evaluate the predictive accuracy of the target model and the interpretative accuracy of the MID surrogate based on the RMSE between the `test` and model prediction or the two predictions, respectively.

```{r utils}
# define utility functions for the following chunks
effect_plots <- function(object) {
  mid.plots(object, terms = paste("x", 1:6, sep = "."))
}

interaction_plot <- function(object) {
  ggmid(object, "x.1:x.2", main.effects = TRUE, theme = "Zissou 1") +
    ggtitle("interaction effect")
}

ice_theme <- color.theme("Zissou 1@seq")
ice_plot <- function(object, data = train[1:200, ]) {
  ggmid(mid.conditional(object, "x.1", data = data),
        var.color = x.2, type = "centered", theme = ice_theme) +
    ggtitle("conditional expectation")
}

importance_plot <- function(object) {
  ggmid(mid.importance(object), "heatmap") +
    ggtitle("feature importance")
}

eval_plot <- function(model, mid, data = test, ...) {
  pred <- get.yhat(model, data, ...)
  pred_mid <- get.yhat(mid, data)
  actual <- test$y
  rmse_vs_test <- rmse(pred, actual)
  rmse_vs_mid <-  rmse(pred, pred_mid)
  ggplot() + scale_color_theme("highlight?accent='#F5191C'") +
    geom_point(aes(x = pred, y = actual, col = "vs test")) +
    geom_point(aes(x = pred, y = pred_mid, col = "vs mid")) +
    geom_abline(slope = 1, intercept = 0, col = "black", lty = 2) +
    labs(x = "model-prediction", y = "mid-prediction / test") +
    annotate(
      "text", family = "serif", size = 3,
      x = min(pred) + diff(range(pred)) / 8,
      y = max(actual) - diff(range(actual) / 8),
      label = sprintf("RMSE\nvs test: %.3f\nvs mid: %.3f",
                      rmse_vs_test, rmse_vs_mid)
    ) + ggtitle("prediction / representation accuracy")
}
```

## Additive Models

### Linear Model

```{r stats_lm}
model <- lm(y ~ ., train)
coef(model)
mid <- interpret(y ~ .^2, mtrain, model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

### Regularized GLM

```{r glmnet_glmnet, message=FALSE}
library(glmnet)
model <- glmnet(x = as.matrix(train[, -11]), y = train[, 11])
# prediction with arbitrarily chosen lambda
mid <- interpret(y ~ .^2, train[, -11], model,
                 pred.args = list(s = model$lambda[9]))
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
evp <- eval_plot(model, mid, data = test[, -11],
                       s = model$lambda[9])
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), evp, nrow = 2)
```

### Generalized Additive Model

```{r gam_gam, message=FALSE}
library(gam)
model <- gam(y ~ s(x.1) + s(x.2) + s(x.3) + s(x.4) + s(x.5) +
             s(x.6) + s(x.7) + s(x.8) + s(x.9) + s(x.10),
             family = gaussian, data = train)
mid <- interpret(y ~ .^2, train, model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

## Neural Network

### Single Hidden Layer Network

```{r nnet_nnet}
library(nnet)
set.seed(42)
model <- nnet(y ~ ., train, size = 5, linout = TRUE, maxit = 1e3, trace = FALSE)
mid <- interpret(y ~ .^2, mtrain, model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

## Support Vector Machine

### RBF Kernel SVM

```{r e1071_svm}
library(e1071)
model <- svm(y ~ ., train, kernel = "radial")
mid <- interpret(y ~ .^2, mtrain, model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

## Tree Based Models

### Gradient Boosting Trees

```{r xgboost_xgboost}
library(xgboost)
params <- list(eta = .1, subsample = .7, max_depth = 5)
set.seed(42)
model <- xgboost(as.matrix(train[, -11]), train[, 11], nrounds = 100,
                 params = params, verbose = 0)
mid <- interpret(y ~ .^2, as.matrix(mtrain), model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
evp <- eval_plot(model, mid, as.matrix(test[, -11]))
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), evp, nrow = 2)
```

### Random Forest

```{r ranger_ranger}
library(ranger)
set.seed(42)
model <- ranger(y ~ ., train, mtry = 5)
mid <- interpret(y ~ .^2, mtrain, model)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

### Decision Tree

```{r rpart_rpart}
library(rpart)
model <- rpart(y ~ ., train)
# create encoding frames for CART
frm <- cbind(model$frame, labels(model, collapse = FALSE))
print(t(frm[frm$var != "<leaf>", c("var", "ltemp")]))
frames <- lapply(mtrain, range)
frames$x.1 <- c(frames$x.1, .2100, .3184)
frames$x.2 <- c(frames$x.2, .3110, .2653, .4782)
frames$x.4 <- c(frames$x.4, .5579, .2953, .8843, .8413)
frames$x.5 <- c(frames$x.5, .5849, .2486)
mid <- interpret(y ~ .^2, mtrain, model, type = 0, frames = frames)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```

## Other Modes

### Predictive MID

```{r mid_interpret}
model <- mid <- interpret(y ~ .^2, train, lambda = .2)
pred <- pred_mid <- predict(mid, test)
print(mid)
grid.arrange(grobs = effect_plots(mid), nrow = 2L)
grid.arrange(interaction_plot(mid), importance_plot(mid),
             ice_plot(mid), eval_plot(model, mid), nrow = 2)
```
